Review: Adaptive Introgression Expanded the Genetic Base of Crops during post-Domestication Spread

### 2017-07-06 paper.tex
edits to barley case study

In this review, we will: 1) briefly describe recently developed methods for detecting adaptive introgression and provide a summary of how they can be applied to detect crop-wild introgression, 2) present case studies suggesting wild-to-crop introgression has conferred local adaptation, 3) consider how adaptive introgression bears upon fundamental questions of domestication, and 4) describe future advances in both basic and applied genetics that can be made through the study of introgression in agroecosystems.

@@

The relationship between maize (\emph{Zea mays} ssp. \emph{mays}) and the teosinte \emph{Zea mays} ssp. \emph{mexicana} (hereafter referred to as \emph{mexicana}) offers a prime case study of adaptive wild-to-crop introgression.
Maize was domesticated from \emph{Zea mays} ssp. \emph{parviglumis} (hereafter referred to as \emph{parviglumis})  approximately 9,000 years ago in the lowlands of the Balsas River Valley in Mexico \cite{matsuoka2002single}.
From this domestication center, maize spread into the highlands of the Mexican Central Plateau, where it came into sympatry with \emph{mexicana}.
Introgression from \emph{mexicana} to maize in the Central Plateau has long been reported based on both morphological \cite {wilkes1977, lauter2004, doebley1984} and molecular \cite{matsuoka2002single, vanHeerwaarden2011, doebley1987, warburton2011, fukunaga2005} data.
However, \citep{hufford2013} first localized \emph{mexicana} introgression to chromosomal regions and provided evidence that it was likely adaptive.
%The following is not fully clear.  Cut or revise.
The authors identified nine genomic regions in several maize populations which consistently showed evidence of \emph{mexicana} introgression based on chromosome painting using both HAPMIX and the linkage model of STRUCTURE (Figure \ref{fig:introgressionMaize}).
These introgressed segments showed low diversity and overlapped QTL that had previously been found to control anthocyanin content and leaf macrohairs \cite{lauter2004}, traits known to be adaptive at high elevation.
In a growth chamber experiment, the authors demonstrated that maize populations with \emph{mexicana} introgression showed greater plant height (a proxy for fitness) under highland environmental conditions than populations that lacked introgression.
Height differences were not detected under lowland conditions.
%Among the nine regions, three span the centromeres of chromosomes 5, 6, and 10, and one is located in the inversion polymorphism on chromosome 4, suggesting a significant role of genome structures restricting recombination in adaptive introgression.
%In summary, the introgressed alleles/haplotypes from \emph{mexicana} to maize conferred adaptation to highland habitats when maize migrated from Mexican lowlands.

@@

%Domesticated and wild barley belong to the same species, \emph{Hordeum vulgare}, and are capable of producing viable offpspring via hybridization \cite{von1995ecographical}.
Barley (\emph{Hordeum vulgare} subsp. \emph{vulgare}) was domesticated at least twice roughly 8,000 to 10,000 BP: once from the wild subsp. \emph{spontaneum} in the Fertile Crescent and once from subsp. \emph{spontaneum} var. \emph{agriocrithon} in Tibet \cite{takahashi1955origin, badr2000origin, oka2012origin, azhaguvel2007phylogenetic, haberer2015barley, ren2013tibet, dai2012tibet}.
\gmj{While morrell2007 cites the Zargos mountains as a center of domestication, this paper is cited by ren2013tibet, which states that the findings of morrell2007genetic are just that a second origin exists east of the Fertile Crescent. It is cited by dai2012tibet and the poets paper, but little is said of its mention of the zargos mountains.  I think that the mention of the Zargos mountains was just as a dividing line between the two spheres of influence between the two domestication centers.  }
Presently, the distribution of subsp \emph{spontaneum} stretches from the eastern Mediterranean through the Middle-East to west-central Asia spanning clines in temperature, precipitation, soil type, and altitude \cite{nevo2010drought}.
Barley/\emph{spontaneum} hybrids are known to be fertile and are found spontaneously when wild and domesticated barleys co-occur.
Introgression between wild and domesticated barleys appears to be frequent \cite{dai2012tibet}at times occuring over distances of more than a kilometer \cite{hillman2001new}.

@@

The authors of \cite{Poets2015} investigated the contribution of wild barleys to domesticated barley landraces from either center of domestication, looking specifically for patterns of contribution across regions of the genome and across geographical ranges.
Broadly, these authors found several lines of evidence that indicate that introgression from many wild populations of barley may have been important in the dispersal and adaptation of domesticated barleys.
Barley landraces show greater genetic contributions from proximate wild relatives than from distant ones, suggesting that these contributions may be locally-adaptive.
Furthermore, genomic segments with greater contribution from a wild population than the average are more likley to contain regionally-unique, locally-adaptive alleles, and unequal genetic contributions at specific genomic regions to individuals within landraces may suggest that adaptive alleles from diverse geographical areas may have been combined.
\gmj{I'll be honest, I don't actually follow the reasoning in the second half of the above sentence.  It is a point made in the Poets paper.  I think my confusion arises from their specific "ancestry" term they define.  I'll have to return to this at another point.}
Low linkage disequilibrium and small blocks of identity by state indicate that these introgressed regions are old, perhaps dating back to the beginning of barley domestication.

Genomic evidence points to a history of wild-crop introgression in the dispersal of barleys beyond their centers of domestication, but for the most part, the identity of the alleles critical to these adaptations have yet to be uncovered.
One exception comes from the authors of \cite{dai2012tibet}, who hypothesize that alleles for cold stress tolerance and spring habit \gmj{"spring habit" are the words from the paper, but they're not defined.  Does this mean that it grows more quickly in the spring?} from wild Tibetan 2-rowed barley allowed the barley to spread to the cold northern regions of Europe.
Wild-domesticate breeding experiments have shown that wild barleys have alleles for several important fitness and agronomic phenotypes, including powdery mildew resistance, brittleness, flowering time, plant height, lodging, and yield \cite{dreiseitl2017heterogeneity,von2006ab,handley1994chromosome}, and these individually or in combination may have been facilitators of the dispersal of domesticated barley.

>>>>

In this review, we will: 1) briefly describe recently developed methods for detecting adaptive introgression and provide a summary of how they can be applied to detect crop-wild introgression, 2) present case studies suggesting wild-to-crop introgression has conferred local adaptation, 3) consider how introgression bears upon fundamental questions of domestication, and 4) describe future advances in both basic and applied genetics that can be made through the study of introgression in agroecosystems.

@@

The relationship between maize (\emph{Zea mays} ssp. \emph{mays}) and the teosinte \emph{Zea mays} ssp. \emph{mexicana} (hereafter referred to as \emph{mexicana}) offers a prime case study of adaptive wild-to-crop introgression.
Maize was domesticated from \emph{Zea mays} ssp. \emph{parviglumis} (hereafter referred to as \emph{parviglumis})  approximately 9,000 years ago in the lowlands of the Balsas River Valley in Mexico \cite{matsuoka2002single}.
From this domestication center, maize spread into the highlands of the Mexican Central Plateau, where it came into sympatry with \emph{mexicana}.
Introgression from \emph{mexicana} to maize in the Central Plateau has long been reported based on both morphological \cite {wilkes1977, lauter2004, doebley1984} and molecular \cite{matsuoka2002single, vanHeerwaarden2011, doebley1987, warburton2011, fukunaga2005} data.
However, \citep{hufford2013} first localized \emph{mexicana} introgression to chromosomal regions and provided evidence that it was likely adaptive.
%The following is not fully clear.  Cut or revise.
The authors identified nine genomic regions in several maize populations which consistently showed evidence of \emph{mexicana} introgression based on chromosome painting using both HAPMIX and the linkage model of STRUCTURE (Figure \ref{fig:introgressionMaize}).
These introgressed segments overlapped QTL that had previously been found to control anthocyanin content and leaf macrohairs \cite{lauter2004}, traits known to be adaptive at high elevation.
In a growth chamber experiment, the authors demonstrated that maize populations with \emph{mexicana} introgression showed greater plant height (a proxy for fitness) under highland environmental conditions than populations that lacked introgression.
Height differences were not detected under lowland conditions.
%Among the nine regions, three span the centromeres of chromosomes 5, 6, and 10, and one is located in the inversion polymorphism on chromosome 4, suggesting a significant role of genome structures restricting recombination in adaptive introgression.
%In summary, the introgressed alleles/haplotypes from \emph{mexicana} to maize conferred adaptation to highland habitats when maize migrated from Mexican lowlands.

@@

%Domesticated and wild barley belong to the same species, \emph{Hordeum vulgare}, and are capable of producing viable offpspring via hybridization \cite{von1995ecographical}.
Barley (\emph{Hordeum vulgare} subsp. \emph{vulgare}) was domesticated at least twice roughly 8,000 to 10,000 BP: once from the wild subsp. \emph{spontaneum} in the Fertile Crescent and once from subsp. \emph{spontaneum} var. \emph{agriocrithon} in Tibet \cite{takahashi1955origin, badr2000origin, oka2012origin, azhaguvel2007phylogenetic, haberer2015barley, ren2013tibet, dai2012tibet}.
%\gmj{While morrell2007 cites the Zargos mountains as a center of domestication, this paper is cited by ren2013tibet, which states that the findings of morrell2007genetic are just that a second origin exists east of the Fertile Crescent. It is cited by dai2012tibet and the poets paper, but little is said of its mention of the zargos mountains.  I think that the mention of the Zargos mountains was just as a dividing line between the two spheres of influence between the two domestication centers.  }
Presently, the distribution of subsp \emph{spontaneum} stretches from the eastern Mediterranean through the Middle-East to west-central Asia spanning clines in temperature, precipitation, soil type, and altitude \cite{nevo2010drought}.
Barley/\emph{spontaneum} hybrids are fertile and found spontaneously when wild and domesticated barley co-occur.
Introgression between wild and domesticated barley is frequent \cite{dai2012tibet}, at times occurring over distances greater than a kilometer \cite{hillman2001new}.

@@

Poets and co-authors \cite{Poets2015} investigated the range-wide contribution of wild barley to landraces, assessing both genome-wide and geographical patterns.
This study identified several lines of evidence consistent with wild introgression aiding the dispersal and adaptation of domesticated barley.
Genomic regions of shared ancestry were detected linking individual landraces to a number of wild relative populations, suggesting landraces may have received wild introgression on a continual basis during post-domestication expansion.
However, barley landraces showed an excess of ancestry from nearby wild relatives, indicating a prevalence of local and potentially adaptive gene flow.
Low linkage disequilibrium and small blocks of identity by state indicated even these locally introgressed regions are old, perhaps dating back to the early expansion of barley following domestication.
While these results are suggestive, wild barley haplotypes have yet to be definitely linked to specific local adaptations in landraces.

%Genomic evidence points to a history of wild-to-crop introgression during dispersal of barley beyond their centers of domestication, but for the most part, the identity of the alleles critical to these adaptations have yet to be uncovered.
%One exception comes from the authors of \cite{dai2012tibet}, who hypothesize that alleles for cold stress tolerance and spring habit \gmj{"spring habit" are the words from the paper, but they're not defined.  Does this mean that it grows more quickly in the spring?} from wild Tibetan 2-rowed barley allowed the barley to spread to the cold northern regions of Europe.
%Wild-domesticate breeding experiments have shown that wild barleys have alleles for several important fitness and agronomic phenotypes, including powdery mildew resistance, brittleness, flowering time, plant height, lodging, and yield \cite{dreiseitl2017heterogeneity,von2006ab,handley1994chromosome}, and these individually or in combination may have been facilitators of the dispersal of domesticated barley.
